---
title: "Docker and Traefik"
subtitle: "Notes by a frustrated newbie"
author: Ott Toomet
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    highlight: tango
    df_print: kable
    fig_caption: false
    toc: true
    toc_float:
      smooth_scroll: false
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE, cache=FALSE}
options(tibble.width=60, tibble.print_max=7, tibble.print_min=4)
knitr::opts_knit$set(aliases=c(h="fig.height"))
knitr::opts_chunk$set(fig.height=100/25.4, 
                      cache=FALSE, 
                      message=FALSE)
library(magrittr)
```

# Introduction

I have next to zero experience with web development, however, for a team
project I agreed to build a REST API for a web app.  Getting the basic
API up and running (using [R/plumber](https://www.rplumber.io/)) was fairly straightforward.  What
turned to a substantial task was to add https security to it.

As plumber does not implement https itself, I had to rely on some sort
of frontend.  I decided to go for [traefik](https://traefik.io/),
apparently a very popular reverse proxy framework, and set everything
up using docker services.
This is probably an easy and quick task for experienced network
people, but for a newbie like me it took a few days of trials and errors and frutstration.  Hope
others will find some of my experience useful.


# How did we get there

In order to automatically extract information from legal documents, we need
high-quality training data.  However, creating it may be a slow and tedious task.  We developed a
web app, based on [swipes for
science](https://swipesforscience.org/) using javascript/axios
framework that queries data from the server, and pre-fills the fields for
faster and less error-prone manual work.  The app gets its data from
the server.  This is why we need the API.

The server is mostly
locked down, but has port 8080 open for the API.  I do not have full
root access to the server.


# Plumber

I chose R and plumber for two reasons: I know R very well, and plumber
seems to be easy for setting up simple API-s.  Indeed, this was the
case.  So far I have had only minimal issues with the plumber side of
the API, and this has been mostly related to my network ignorance.
Just two notes: in order to make you API accessible from outside, you
have to

* allow CORS by setting something like
```{r eval=FALSE}
cors <- function(req, res) {
   res$setHeader("Access-Control-Allow-Origin", "*")
   plumber::forward()
}
```
* bind it to 0.0.0.0 by something like
```{r eval=FALSE}
pr <- plumber::plumb('api.R')
pr$run(port=args$port,  host="0.0.0.0")
```

(Unfortunately I still have only a vague what all this means...)


# docker

Installing docker was trivial.  Also, the documentation needed to get my API
up and running in the docker was sufficiently available and soon I was
able to create my own container with API in it.  My dockerfile looks
like this:

```{dockerfile}
FROM r-base
LABEL description="run the API server using local port 7887"
RUN apt-get update
RUN apt-get install --no-install-recommends --yes imagemagick
RUN ["Rscript", "-e", "install.packages(c('data.table', 'plumber', 'magrittr', 'base64enc', 'argparser', 'log4r', 'R.utils'))"]
WORKDIR api
COPY api/api.R api/startServer.R bin/conf.R bin/utils.R ./
VOLUME ["/data", "/project", "/logs"]
EXPOSE 7887
CMD ["Rscript", "startServer.R", "--port", "7887"]
```

As it was not hard to create the docker container, I explain these
steps only briefly.

* **FROM** imports the base container.  This is essentially your operating
  system your stuff will run on later.  _r-base_ is a stripped-down
  ubuntu with the latest R installed.
* I also install imagemagick as my API converts images.  As _r-base_
  is ubuntu-based, you install packages using the usual apt family of
  commands, just preceded with **RUN** to run a linux command. 
* Next, I install all necessary R packages.
* **COPY** copies the necessary code files into current **WORKDIR**,
  the directory where the code will be run later
* **VOLUME** declares external mounts, external in the sense that
  these link to the world outside of the container.  These are mounted
  later.
* **EXPOSE** tells the container that the port 7887 should be
  accessible from outside of it.  Note that you still have to map it
  somewhere on the host or another container.
* and finally, **CMD** is the command to start the API.


# Traefik: where I got stuck

Plumber does not implement https/TLS security.  I agree that
mission-critical components, such as security, is better to leave for
dedicated software.  Traefik seems to be a popular modern choice in
this regard.

Unfortunately, traefik's documentation is much sparser, and loaded
with network terminology (well, what else can you expect from a
network proxy...).  There are number of easy intros available but my
case has a few notable differences: first, I cannot use the port 8080 on server
for traefik webUI, at least not directly as this port is the only
one open for the API.  Second, I do not want to keep http open, only
https.  Third, I do not want to use _LetsEncrypt_ TLS certificate as I
have a dedicated certificate for our workgroup.  Finally, I also want
to pull up both API and traefik (and potentially more services)
together through `docker stack deploy`.  I was unable to find a good
introduction that checked all these boxes.

## docker stack configuration `docker-compose.yml`

For `docker stack deploy` we need `docker-compose.yml` configuration
file.  This lists all the _services_ to be run (here _traefik_ and
_api_) and defines how they will be initiated, how these interact, and
how the volumes on the host and containers are linked.
This is the current version:

```{yaml}
version: '3.5'

services:
  api:
    image: otoomet/api
    # ports:
    #   - ":7887"
    #   if I uncomment this, the container's 7887 will be exposed on
    #   the hosts 8080 bypassing traefik
    deploy:
      labels:
        - traefik.backend=api
        - "traefik.frontend.rule=Host:server.example.com,localhost;PathPrefixStrip:/swipes"
        # works both on tess and localhost
        - traefik.port=7887
    volumes:
      - $DATAROOT:/data:ro
      - $PROJECTDIR:/project
      - $HOME/logs:/logs
    networks:
      - traefik-public
  traefik:
    image: traefik
    ports:
      - "8079:8079"
      # The Web UI, keep 8080 clear for API
      - "8080:8080"
      # API entry
    volumes:
      - /var/run/docker.sock:/var/run/docker.sock
      # So that Traefik can listen to the Docker events
      - $CODEDIR/api/traefik.toml:/etc/traefik/traefik.toml
      - $LOGSDIR:/logs
      - $KEYFILE:/cert/swipes.key:ro
      - $CERTFILE:/cert/swipes.crt:ro
    deploy:
      labels:
        - traefik.enable=true
        # needed for traefik UI
    networks:
      - traefik-public
networks:
  traefik-public:
    external: true
    name: traefik-public
```

It contains quite a few entries I discovered through trial and error,
sometimes it took me hours of experimenting.  Notes:

* when declaring the services, do not declare ports as `8080:7887`.
  This would expose the service container's 7887 as hosts 8080
  bypassing traefik and TLS security.
* `deploy` and `labels`.  The documentation indicates that `labels`
  for services should be under `deploy`.  However, I was able to get
  API up (but not traefik GUI) without the `deploy` section by
  declaring `labels` directly under `api` adn `traefik`.  Even more,
  moving certain labels under `deploy` did not work.  I still don't
  know what's going on but I suspect that if you define labels under
  `deploy`, you have to define all labels there.
* `traefik.backend=api`.  This is tells traefik to connect backend
  _api_ in traefik.toml configuration file to the `api` service here.
  If not specified, **traefik fails to connect** to `api`!
*
  `traefik.frontend.rule` is necessary to tell traefik which requests
  that hit it should be forwarded to `api`.  We have three rules here:
    - `Host:server.example.com` means that the requests must have host
      as `server.example.com`.  
    - Ditto with `localhost`, for testing
      purposes.  Note the comma inbetween, this means that both
      `server.example.com` and `localhost` qualify as queries to be
      redirected. 
	- finally, `PathPrefixStrip:/swipes` tells traefik that not
      everything going to these hosts should be directed to the api
      service, only queries in the form
      `server.example.com/swipes/...` and `localhost/swipes/...`.  And
      the `Strip`-part tells that when forwarded, the `swipes`-part
      must be stripped from the query.
	  
* Traefik documentation also shows how to set the rules in
  `traefik.toml` file but I haven't been able to get those rules
  loaded.  So I keep my rule here.  
  `traefik.frontend.rule` is a must.  Otherwise traefik simply does
  not forward anything to the API.
* finally, `traefik.port=7887` tells traefik to forward the relevant
  requests to port 7887 to the api-s container.  This is also a must.
* `volumes`: this is the way to connect your containers to the outside
  world in the sense of the host file system.  `docker-compose.yml`
  happily accepts environment variables, so I have a separate
  cofiguration script that sets these to the relevant folders/files.
  This also allows me to easily change between the production and mockup
  setup. 
* finally, `networks`.  Here I use a custom separately created network
  to keep my stack running.  I don't think it is necessary, if not
  specified, the automatically generated network should do.  But if
  frequently restarting during tests, the automatically generated may
  fail to start sometimes, so it is a nice-to-have thing.  You can
  create it with the command `docker network create -d overlay traefik-pubic`.  The _overlay_-driver is necessary.  Note also that
  if you create the external network, you have to list it in the
  separate network section (here at the end) as external.

This was about the API service.  Next _traefik_.

* `ports`: I want to expose both the API port 8080 and 8079 for the
  UI.  (the latter is normally blocked from outside but still
  accessible on server).
* `volumes`: as above, these are the channels how the containered
  service links to the host file system.  Two comments:
  `/var/run/docker.sock` entry is necessary for traefik to be able to
  talk to the docker daemon; second note that `$KEYFILE` and
  `$CERTFILE` are files, not folders.
* `deploy`-`labels`: `traefik.enable=true` is needed to get the UI
  up.
  

## traefik configuration `traefik.toml`

This is a place where we should be able to include most of the traefik
configuration.  However,  I have not had success with all options.

```{toml}
defaultEntryPoints = ["https"]

[docker]
  endpoint = "unix:///var/run/docker.sock"
  swarmMode = true
  
[web]
  address = ":8079"

logLevel = "INFO"

[entryPoints]
  [entryPoints.https]
    address = ":8080"
    [entryPoints.https.tls]
      [[entryPoints.https.tls.certificates]]
      certFile = "/cert/swipes.crt"
      keyFile = "/cert/swipes.key"

[frontends]
  [frontends.swipes]
  backend = "api"
```

Comments:

* `defaultEntryPoints = ["https"]`: sorry, I don't know what it does.
  I think it simply tells traefik where to direct the requests, and
  has nothing to do with https security.
* `[docker]` `endpoint` is a must and must be connected to the docker
  daemon socket in the _docker-compose.yml_ file.
* `[docker]` `swarmMode` not quite sure what it does--I had the API
  working without it, just I was unable to get traefik UI up.  I
  suspect it is related to whether the labels are listed in _deploy_
  section in the _docker-compose.yml_ file.
* `[web]` `address` is the graphical UI address.  It is normally 8080
  but I need another port, as explained above.
* `logLevel="INFO"`: it is obvious.  Well, it should be.  So far I
  have not been able to log anything else besides error messages.  So
  I don't know what is wrong here.
* `entryPoints.https`:
    - address = ":8080" means this entrypoint should listen to 8080.
      This is the only port open on my server.
    - the next entries point to the files of TLS certificates.  These
      files were mounted under volumes in the _docker-compose.yml_
      above.  All these entries together do the magick of https for
      port 8080.

* finally, `frontends` lists the different services accessible from
  outside, and the corresponding `backend` tells which backend must be
  invoked when the corresponding request arrives.  These must
  correspond to the _traefik.backend_ and _trafik.frontend.rule_
  entries in the _docker-compose.yml_ file.
  
All this is probably trivial for a network guru, and seems pretty
straightforward even when just reading it here.  But it is not at all
straighforward when confronted with traefik documentation with little
network knowledge.


# HTTPS and TLS certificates

Certificates are needed for https encryption.  Major websites use
certificates approved by common authorities, public keys of which are
built into common software tools.  But you can easily create a
self-signed one using `openssl`.  These come in pairs, essentially
public and private key, and if using a self-signed one, you have to
provide the public key to your network software.  Otherwise you may
get more or less explanatory network errors.

Note that many common programs assume the certificates have file
extension _key_, _crt_, or _pem_.

Note that you can "ping" the https-protected services with http.  You
get a binary response `15 03 01 00 02 02 0a`, telling you that the TLS
security layer did not understand the HTTP request.
